\Large Generative Dynamics of Supreme Court Citations

### 2020-04-20 Tex/paper.tex
Updated Section 3 The citation ERGM

 \section{The Citation Temporal Exponential Random Graph Model}
{\color{blue} This section needs to be completely revised. Model is not a TERGM anymore. - CS} 
We develop a methodology that can be used to jointly test for the effects of covariates on citations---as have been studied in prior research, and test for dependence effects, as we hypothesize above. To accomplish this, we extend a model that has been developed to jointly model covariate and dependence effects in longitudinal network data---the temporal exponential random graph model (TERGM) \citep[e.g.,][]{cranmer2012complex,clark2013multimember, masket2015polarization, graif2017neighborhood}. The TERGM, as it is currently implemented \citep{leifeld2017temporal,  tergm}, is designed to model longitudinal network data in which ties among a relatively stable set of nodes are modeled at multiple time points (e.g., the countries that are and are not defensively allied at time $t$, the legislators who have and have not cosponsored each others' legislation in the current session).  The longitudinal network structures for which TERGMs are currently designed are insufficient to account for the structure of citation networks.

\subsection{c-TERGM definition}

We develop the citation TERGM (c-TERGM), to account for the structural constraints that apply to the network of Supreme Court citations. These structural constraints amount to three departures from the structure of longitudinal networks for which TERGMs are currently designed. First, the citation network is partially acyclic. If two cases are decided during the same term, they can cite each other, forming a mutual edge (or two-cycle). However, if case $i$ is decided before case $j$, case $j$ can cite case $i$, but case $i$ cannot cite case $j$. Second, new edges can be created over time, but cannot be eliminated. Unlike in, e.g., an alliance network, in which two countries can dissolve an alliance, once a citation exists in a citation network it cannot be dissolved. Third, the set of nodes in the network must increase for new edges to be created. In conventional TERGMs, the number of nodes in the network can increase or decrease in each time period, and is typically stable over time. In a citation network, new edges (and non-edges) are introduced over time via the introduction of new nodes. The structure of the citation network is depicted in Figure \ref{fig:ctergm}---a hypothetical citation network established over three time periods, with three cases decided in each time period. We denote $C_t$ to be the set of citation and non-citations added to the network at time $t$ (i.e., via the addition of three cases), $C_{ <t}$ to be the citations among cases decided before time $t$, and $C_{ \leq t}$ to denote the entire set of citations and non-citations on which $C_t$ can depend through the c-TERGM specification.

@@

The likelihood function of the c-TERGM is given by
\begin{equation}
l(\bm{\theta},C_{\leq T}) =  \prod_{t=1}^T \frac{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right] }{ \sum_{C_t^* \in \mathcal{C}_t} \exp \left[ {\bm{\theta}'\bm{h}(C^*_{t},C_{<t}) }\right]  },
\end{equation}
where $\bm{\theta}$ is a vector of real-valued parameters, and $\bm{h}(C_{t},C_{<t})$ is a vector of scalar-valued functions that each quantify a feature of the citation network (e.g., the relationship between citation ties and a case attribute, the number of mutual edges in the citation network). $\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right]$ is a positive weight that is proportional to the probability of observing any particular form of the citations and non-citations added to the network at time $t$.  The denominator of Equation 1 represents a normalizing constant, in which the positive weight is summed over all possible configurations of $C_{t}$, from the network in which the cases added to the network at time $t$ send no citations at all, to the network in which the cases added at time $t$ cite every possible case, and everything in between. The probability of observing the citation network up to time $T$---the final time point in the dataset---is given by the product over the probability of adding the citations at each time point $t$, conditional on the citations that were added to the network previously.

Though the likelihood function of the c-TERGM may appear quite different from that of statistical models conventionally used in political science, analysis of the conditional probability of a single citation from case $i$ to case $j$ reveals that we can interpret the parameters similar to logistic regression coefficients. $$ P(C_{ij,t} = 1 | C_{-ij,t}, C_{ < t}) = \frac{\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right]}{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right] + \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 0) } \right]}, $$ $$ \text{~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~} = \frac{1}{ 1 + \exp \left[ - {\bm{\theta}'\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)} \right]}, $$
where $C_{ij,t} = 1$ indicates that case $i$ cites case $j$, $C_{ij,t} = 0$ indicates that case $i$ does not cite case $j$, $C_{-ij,t}$ is the observed elements of $C_{t}$ except $C_{ij,t}$, and $\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)$ is the change in $\bm{h}(C_{t},C_{<t})$ that results from toggling $C_{ij,t} = 0$ to $C_{ij,t} = 1$. This re-arrangement illustrates that the parameters can be interpreted in terms of the change in the log odds of a citation from $i$ to $j$ given a one-unit increase in the corresponding element of $\bm{h}$, conditional on the other citations observed in the network. For example, if the value of $\theta$ corresponding to an element of $\bm{h}$ that counts the number of mutual edges in the network is 0.5, then the log odds of observing $C_{ij,t} = 1$ increase by 0.5 if case $i$ is cited by case $j$ (as compared to the configuration in which case $j$ does not cite case $i$). The logit form conditional probability is not unique to c-TERGM. It is well known for the ERGM family \citep{goodreau2009birds}.

We should note that there is nothing about the c-TERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, the term for that dynamic would indicate that there is an increase in the log odds that case $i$ cites case $j$ with each other case that has cited case $j$.  That finding could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case---as in the Baker v. Carr passage cited above. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-TERGM can be specified to account for the dependence in the data.

\subsection{c-TERGM Estimation}

The normalizing constant in Equation 1 is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure \ref{fig:ctergm}---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that already includes thousands of previous cases. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with the c-TERGM. Methods of Monte Carlo approximation are well established in the ERGM literature, \citep{hunter2006inference,van2009framework,hummel2012improving}, but with a number of nodes on the order of $10,000$, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. However, because the network is large, we can rely on a fast estimation method---maximum pseudo-likelihood estimation (MPLE), which is consistent for discrete exponential family models \citep{nguyen2017near}, and has long been used with ERGM \citep{strauss1990pseudolikelihood}. The MPLE is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated above, assume a logit form.

One challenge that arises with the use of MPLE is that the usual estimate of the asymptotic covariance matrix (i.e., inverting the negated Hessian of the objective function at the maximum), results in standard errors that are biased downward. This leads to a high Type-1 error rate in hypothesis testing with MPLE if conventional standard errors are used \citep{van2009framework}. A form of non-parametric cluster bootstrap results in confidence intervals with correct coverage probability with MPLE \citep{desmarais2012statistical,desmarais2010consistent}. In the case of c-TERGM, the clusters are defined by $t$---the time points in which the cases are decided, and $T$ time-specific citation matrices $C_{t}$ are re-sampled with replacement to form a non-parametric cluster bootstrap sample. Confidence intervals are defined by the quantiles in the empirical distributions of estimates that result from re-estimating the MPLE on non-parametric cluster bootstrap samples.

>>>>

\section{The Citation Exponential Random Graph Model}
We develop a methodology that can be used to jointly test for the effects of covariates on citations---as have been studied in prior research, and test for dependence effects, as we hypothesize above. To accomplish this, we extend a model that has been developed to jointly model covariate and dependence effects in network data---the exponential random graph model (ERGM) \citep[e.g.,][]{cranmer2012complex,clark2013multimember, masket2015polarization, graif2017neighborhood} {\color{blue} Citations need to be adjsuted to ERGM, not TERGM- CS}. %The ERGM, as it is currently implemented \citep{leifeld2017temporal,  tergm}, is designed to model longitudinal network data in which ties among a relatively stable set of nodes are modeled at multiple time points (e.g., the countries that are and are not defensively allied at time $t$, the legislators who have and have not cosponsored each others' legislation in the current session).  
The network structures for which ERGMs are currently designed are insufficient to account for the structure of citation networks.

\subsection{c-ERGM definition}

We develop the citation ERGM (c-ERGM), to account for the structural constraints that apply to the network of Supreme Court citations. These structural constraints amount to three departures from the structure of networks for which ERGMs are currently designed. First, the citation network is partially acyclic. If two cases are decided during the same term, they can cite each other, forming a mutual edge (or two-cycle). However, if case $i$ is decided before case $j$, case $j$ can cite case $i$, but case $i$ cannot cite case $j$. Second, new edges can be created over time, but cannot be eliminated. Unlike in, e.g., an alliance network, in which two countries can dissolve an alliance, once a citation exists in a citation network it cannot be dissolved. Third, the set of nodes in the network must increase for new edges to be created. In conventional ERGMs, the number of nodes in the network can increase or decrease in each time period, and is typically stable over time. In a citation network, new edges (and non-edges) are introduced over time via the introduction of new nodes. The structure of the citation network is depicted in Figure \ref{fig:ctergm}---a hypothetical citation network established over three time periods, with three cases decided in each time period. We denote $C_t$ to be the set of citation and non-citations added to the network at time $t$ (i.e., via the addition of three cases), $C_{ <t}$ to be the citations among cases decided before time $t$, and $C_{ \leq t}$ to denote the entire set of citations and non-citations on which $C_t$ can depend through the c-ERGM specification.

@@

The likelihood function of the c-ERGM is given by
\begin{equation}
l(\bm{\theta},C_{\leq t}) =  \frac{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right] }{ \sum_{C_t^* \in \mathcal{C}_t} \exp \left[ {\bm{\theta}'\bm{h}(C^*_{t},C_{<t}) }\right]  },
\end{equation}
where $\bm{\theta}$ is a vector of real-valued parameters, and $\bm{h}(C_{t},C_{<t})$ is a vector of scalar-valued functions that each quantify a feature of the citation network (e.g., the relationship between citation ties and a case attribute, the number of mutual edges in the citation network). $\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right]$ is a positive weight that is proportional to the probability of observing any particular form of the citations and non-citations added to the network at time $t$.  The denominator of Equation 1 represents a normalizing constant, in which the positive weight is summed over all possible configurations of $C_{t}$, from the network in which the cases added to the network at time $t$ send no citations at all, to the network in which the cases added at time $t$ cite every possible case, and everything in between. %The probability of observing the citation network up to time $T$---the final time point in the dataset---is given by the product over the probability of adding the citations at each time point $t$, conditional on the citations that were added to the network previously.

Though the likelihood function of the c-ERGM may appear quite different from that of statistical models conventionally used in political science, analysis of the conditional probability of a single citation from case $i$ to case $j$ reveals that we can interpret the parameters similar to logistic regression coefficients. $$ P(C_{ij,t} = 1 | C_{-ij,t}, C_{ < t}) = \frac{\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right]}{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right] + \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 0) } \right]}, $$ $$ \text{~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~} = \frac{1}{ 1 + \exp \left[ - {\bm{\theta}'\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)} \right]}, $$
where $C_{ij,t} = 1$ indicates that case $i$ cites case $j$, $C_{ij,t} = 0$ indicates that case $i$ does not cite case $j$, $C_{-ij,t}$ is the observed elements of $C_{t}$ except $C_{ij,t}$, and $\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)$ is the change in $\bm{h}(C_{t},C_{<t})$ that results from toggling $C_{ij,t} = 0$ to $C_{ij,t} = 1$. This re-arrangement illustrates that the parameters can be interpreted in terms of the change in the log odds of a citation from $i$ to $j$ given a one-unit increase in the corresponding element of $\bm{h}$, conditional on the other citations observed in the network. For example, if the value of $\theta$ corresponding to an element of $\bm{h}$ that counts the number of mutual edges in the network is 0.5, then the log odds of observing $C_{ij,t} = 1$ increase by 0.5 if case $i$ is cited by case $j$ (as compared to the configuration in which case $j$ does not cite case $i$). The logit form conditional probability is well known for the ERGM family \citep{goodreau2009birds}.

We should note that there is nothing about the c-ERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, the term for that dynamic would indicate that there is an increase in the log odds that case $i$ cites case $j$ with each other case that has cited case $j$.  That finding could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case---as in the Baker v. Carr passage cited above. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-ERGM can be specified to account for the dependence in the data.

\subsection{c-ERGM Estimation}

The normalizing constant in Equation 1 is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure \ref{fig:ctergm}---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that already includes thousands of previous cases. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with the c-ERGM.

The common alternative relates on Monte Carlo methods to approximate the normalizing constant by simulating a large set of networks \citep{hunter2006inference,hummel2012improving}. The resulting estimator, the Monte Carlo MLE (MCMLE), is approximately consistent, meaning that it converges to the MLE as the sample size, i.e., the number of simulated networks, increases. However, one drawback is that with the number of nodes in the Supreme Court citation network being in the order of $10,000$, obtaining the MCMLE is computationally expensive \citep{schmid2017exponential} and the success of the algorithm heavily relies on the starting parameter vector $\theta_0$, which ideally is in the proximity of the MLE \citep{hummel2012improving}. The prevailing choice for $\theta_0$ is the maximum pseudo-likelihood estimation (MPLE) \citep{strauss1990pseudolikelihood}, a fast estimation method that is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated above, assume a logit form. The MPLE is simple to obtain, but does not guarantee a starting value close to the MLE \citep{SchmidHunter2020}.

We estimate the MLE for every term of the c-ERGM between 1950 and 2015. While the network for the 1950 term consists of $1962$ nodes, the size of the network monotonically increases to $10,020$ nodes for the 2015 term, making estimation more challenging as the network grows. The MCMLE of networks up until the 90s was obtainable in a reasonable time frame starting at the MPLE and sampling $10,000$ networks to approximate the normalizing constant. However, the estimation of most networks in the 90s with the MPLE as starting valus was not obtainable in an reasonable time frame anymore. Instead, we fixed the MCMLE of the previous term $t-1$ as starting value to succesfully obtain the MCMLE at time $t$.

But even this approach started to fail for networks at the turn of the millennium. Neither the MPLE nor the MCMLE of previous terms as starting values led to successful estimation, and neither did the Stepping algorithm \citep{hummel2012improving}. 
The MCMLE for these large citation networks was finally obtained by setting the starting value according to an approach introduced by \citet{SchmidHunter2020}. This method is based on the fact that the MLE of exponential family distributions is solely a function of the vector of sufficient statistics $h(C_t, C_{<t})$, meaning that the MLE of two networks $A$ and $B$ is equal if $h(A)=h(B)$. Instead of starting the MCMLE algorithm at the MPLE, \citet{SchmidHunter2020} propose searching for a network $C_t^*$ using simulated annealing \citep{Kirkpatrick83}, with $h(C_t, C_{<t}) = h(C_t^*, C_{<t})$ and a weak dependence structure among unfixed ties. The MPLE of such a network is similar to the MLE, making it an effective starting value for the MCMLE algorithm.

%Methods of Monte Carlo approximation are well established in the ERGM literature, \citep{hunter2006inference,van2009framework,hummel2012improving}, but with a number of nodes on the order of $10,000$, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. However, because the network is large, we can rely on a fast estimation method---maximum pseudo-likelihood estimation (MPLE), which is consistent for discrete exponential family models \citep{nguyen2017near}, and has long been used with ERGM \citep{strauss1990pseudolikelihood}. The MPLE is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated above, assume a logit form.

%One challenge that arises with the use of MPLE is that the usual estimate of the asymptotic covariance matrix (i.e., inverting the negated Hessian of the objective function at the maximum), results in standard errors that are biased downward. This leads to a high Type-1 error rate in hypothesis testing with MPLE if conventional standard errors are used \citep{van2009framework}. A form of non-parametric cluster bootstrap results in confidence intervals with correct coverage probability with MPLE \citep{desmarais2012statistical,desmarais2010consistent}. In the case of c-TERGM, the clusters are defined by $t$---the time points in which the cases are decided, and $T$ time-specific citation matrices $C_{t}$ are re-sampled with replacement to form a non-parametric cluster bootstrap sample. Confidence intervals are defined by the quantiles in the empirical distributions of estimates that result from re-estimating the MPLE on non-parametric cluster bootstrap samples.
